---
title: "Statistical Learning"
author: "ISLR- Done by Prateek Minhas"
date: "2025-10-18"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Ch-2: Statistical Learning 

## Lab

### Basic commands 

```{r}
x=c(1,3,4,5)
x
y=c(4,5,80,-3)
y
length(x)
length(x)+length(y)
x-y
z=x+y

```

#### ls() and Rm()

```{r}
ls()
rm(z)
ls()
rm(list=ls()) #empty out the list
ls()
```

#### matrix() fucntion

```{r}

x = matrix(data=c(1,2,3,4,5,6), nrow=2 , ncol = 3, byrow=TRUE)
x
y = matrix(data=c(1,2,3,4,5,6), nrow=2 , ncol = 3, byrow=FALSE)
y
```

removing the data=, nrow=, ncol= we can also write directly , by default we get byrows = false meaning the columns get filled first

```{r}
z= matrix(c(1,2,3,4,5,6), 3,2)
z
sqrt(x)
y^2
```

**rnorm()** gives random values every time we use it

```{r}
x=rnorm(50)

y=x+rnorm(50,mean=50, sd=.1)
cor(x,y)
```

we use set seed() so that we get same random numbers every time for a particular seed value

```{r}
set.seed(1303)
rnorm(50)
```

```{r}
set.seed(3)
y=rnorm(100)
mean(y)
var(y)
sqrt(var(y))
sd(y)
#as we see variance = standard deviation.sq
```

### Graphics

**plot()**

```{r}
x=rnorm(100)
y=rnorm(100)
plot(x,y, col="green")
plot(x,y,xlab = "this is x axis", ylab = "this is Y axis", main="Plot ox X vs Y")
```

**Seq()**

```{r}
x=seq(1,10)
x
y=1:10
y
z=seq(-pi,pi,length=50)
z
```

**contour() for 3d graphs**

```{r}
y=x
f= outer(x,y, function(x,y) cos(y)/(1+x^2))
contour(x,y,f)
contour(x,y,f , nlevels = 45 , add = T)
fa=(f-t(f))/2
contour(x,y,fa,nlevels = 15)
```

```{r}
image(x,y,fa)
persp(x,y,fa)
persp(x,y,fa, theta = 90, phi=30)
persp(x,y,fa, theta = 60, phi=20)
persp(x,y,fa, theta = 40, phi=15)
```

### Indexing data

```{r}
A= matrix(1:16, 4,4)
A
A[2,3]
```

```{r}
A[c(1,3),c(2,4)]
```

```{r}
A[1:3,2:4]
```

```{r}
A[1:2,]
```

```{r}
A[,1:2]
```

```{r}
A[1,]
```

-ve sign indicates that select all the elements Except those mentioned , it is like a negation

```{r}
A
A[c(1,3),]
A[-c(1,3),]#second and last row 
```

```{r}
A[-c(1,3),-c(1,3,4)]
```

```{r}
dim(A)
```

### Loading Data
