---
title: "Statistical Learning"
author: "ISLR- Done by Prateek Minhas"
date: "2025-10-18"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Ch-2: Statistical Learning

## Lab

### Basic commands

```{r}
x=c(1,3,4,5)
x
y=c(4,5,80,-3)
y
length(x)
length(x)+length(y)
x-y
z=x+y

```

#### ls() and Rm()

```{r}
ls()
rm(z)
ls()
rm(list=ls()) #empty out the list
ls()
```

#### matrix() fucntion

```{r}

x = matrix(data=c(1,2,3,4,5,6), nrow=2 , ncol = 3, byrow=TRUE)
x
y = matrix(data=c(1,2,3,4,5,6), nrow=2 , ncol = 3, byrow=FALSE)
y
```

removing the data=, nrow=, ncol= we can also write directly , by default we get byrows = false meaning the columns get filled first

```{r}
z= matrix(c(1,2,3,4,5,6), 3,2)
z
sqrt(x)
y^2
```

**rnorm()** gives random values every time we use it

```{r}
x=rnorm(50)

y=x+rnorm(50,mean=50, sd=.1)
cor(x,y)
```

we use set seed() so that we get same random numbers every time for a particular seed value

```{r}
set.seed(1303)
rnorm(50)
```

```{r}
set.seed(3)
y=rnorm(100)
mean(y)
var(y)
sqrt(var(y))
sd(y)
#as we see variance = standard deviation.sq
```

### Graphics

**plot()**

```{r}
x=rnorm(100)
y=rnorm(100)
plot(x,y, col="green")
plot(x,y,xlab = "this is x axis", ylab = "this is Y axis", main="Plot ox X vs Y")
```

**Seq()**

```{r}
x=seq(1,10)
x
y=1:10
y
z=seq(-pi,pi,length=50)
z
```

**contour() for 3d graphs**

```{r}
y=x
f= outer(x,y, function(x,y) cos(y)/(1+x^2))
contour(x,y,f)
contour(x,y,f , nlevels = 45 , add = T)
fa=(f-t(f))/2
contour(x,y,fa,nlevels = 15)
```

```{r}
image(x,y,fa)
persp(x,y,fa)
persp(x,y,fa, theta = 90, phi=30)
persp(x,y,fa, theta = 60, phi=20)
persp(x,y,fa, theta = 40, phi=15)
```

### Indexing data

```{r}
A= matrix(1:16, 4,4)
A
A[2,3]
```

```{r}
A[c(1,3),c(2,4)]
```

```{r}
A[1:3,2:4]
```

```{r}
A[1:2,]
```

```{r}
A[,1:2]
```

```{r}
A[1,]
```

-ve sign indicates that select all the elements Except those mentioned , it is like a negation

```{r}
A
A[c(1,3),]
A[-c(1,3),]#second and last row 
```

```{r}
A[-c(1,3),-c(1,3,4)]
```

```{r}
dim(A)
```

### Loading Data

```{r}
Auto = read.table("Auto.data")
#View(Auto)
head(Auto)
```

```{r}
Auto = read.table("Auto.data", na.strings = "?", header = T , stringsAsFactors = T )
head(Auto)
```

```{r}
Auto = read.csv("Auto.csv", na.strings = "?", stringsAsFactors = T)
attach(Auto)
head(Auto)
```

the strings as factors argument tells that if the value is string , treat the variable as qualitative

```{r}
dim(Auto)
```

```{r}
Auto[1:4,]
```

```{r}
Auto= na.omit(Auto)
dim(Auto) #as we see number of rows are reduced
```

```{r}
names(Auto)
```

#### Additional graphical and numerical Summaries

```{r}

plot(cylinders,mpg)
#the cylinders is quantitative
```

The **as.factor() function converts quantitative variables into qualitative** variables.

```{r}
cylinders= as.factor(cylinders)
```

***If the variable plotted on the x-axis is qualitative, then boxplots will automatically be produced by the plot() function.***

```{r}
plot(cylinders, mpg , xlab="cylinderas(qual)", ylab="mpg", col="red", varwidth=T, horizontal=F)
```

```{r}
hist(mpg)
hist(mpg, col=2)
hist(mpg,col=0, breaks=15)
```

The **pairs()** function creates a scatterplotmatrix, i.e. a scatterplot for every pair of variables.

```{r}
pairs(Auto)
pairs(~mpg + displacement + cylinders + acceleration, data=Auto)
```

```{r}
#windows()
plot(horsepower,mpg, col=12)
#identify(horsepower, mpg , name)

```

Use windows() to run the identify interactive function

```{r}
summary(Auto)
```

```{r}
summary(displacement)
```

## APPLIED EXERCISE

8.  This exercise relates to the College data set, which can be found in the file College.csv on the book website. It contains a number of variables for 777 different universities and colleges in the US

<!-- -->

(a) Use the read.csv() function to read the data into R. Call the loaded data college. Make sure that you have the directory set to the correct location for the data.

```{r}
college = read.csv("College.csv" ,stringsAsFactors = T )
attach(college)
View(college)
dim(college)
```

(b) Look at the data using the View() function. You should notice that the first column is just the name of each university. We don’t really want R to treat this as data.

```{r}
rownames(college)= college[,1]
college = college[,-1]
```

(c) 

    i.  Use the summary() function to produce a numerical summary of the variables in the data set.

```{r}
summary(college)
```

```{r}
names(college)
```

ii\. Use the pairs() function to produce a scatterplot matrix of the first ten columns or variables of the data. Recall that you can reference the first ten columns of a matrix A using A[,1:10].

```{r}
pairs(college[,1:10], col=12)
```

iii. Use the plot() function to produce side-by-side boxplots of Outstate versus Private.

```{r}
plot(Private, Outstate)

```

*Here we can see that in average Private Institutes have more out of state tuition*

**iv.** Create a new qualitative variable, called Elite, by binning the Top10perc variable. We are going to divide universities into two groups based on whether or not the proportion of students coming from the top 10% of their high school classes exceeds 50%.

```{r}
if (!("Elite" %in% colnames(college))) {
  Elite = rep("No", nrow(college))
Elite[Top10perc>50]= "Yes"
Elite=as.factor(Elite)
college=data.frame(college,Elite)
}

head(college)
```

Use the summary() function to see how many elite universities there are. Now use the plot() function to produce side-by-side boxplots of Outstate versus Elite.

```{r}
summary(Elite)
```

```{r}
plot(Elite,Outstate, col=12)
```

*So in average elite people have more out of state tuition*

**v.** Use the hist() function to produce some histograms with differing numbers of bins for a few of the quantitative variables. You may find the command par(mfrow = c(2, 2)) useful: it will divide the print window into four regions so that four plots can be made simultaneously. Modifying the arguments to this function will divide the screen in other ways.

```{r}
par(mfrow=c(2,2))
hist(Top10perc)
hist(Top10perc, breaks = 15)
hist(Outstate)
hist(Outstate, breaks=15)
```

**vi. Continue exploring the data, and provide a brief summary of what you discover.**

**9.** This exercise involves the Auto data set studied in the lab. Make sure that the missing values have been removed from the data.
**(a)** Which of the predictors are quantitative, and which are qualitative?

```{r}
head(Auto)
Auto= na.omit(Auto)

```

*Quantitative- mpg, cylinders, displacement, horsepower, weight, acceleration,*

*Qualitative - Year, Origin, name*

(b) What is the range of each quantitative predictor? You can answer this using the range() function.

```{r}
cylinders= as.numeric(as.character(cylinders))
range(mpg)
range(cylinders)
range(displacement)
range(horsepower, na.rm=T)
range(weight)
range(acceleration)
```

(c) What is the mean and standard deviation of each quantitative predictor?

```{r}
cat("MPG Mean:", mean(mpg), "\n")
cat("MPG SD:", sd(mpg), "\n")
cat("displacement Mean:", mean(displacement), "\n")
cat("displacement SD:", sd(displacement), "\n")
cat("weight Mean:", mean(weight), "\n")
cat("weight SD:", sd(weight), "\n")
cat("horsepower Mean:", mean(horsepower), "\n")
cat("horsepower SD:", sd(horsepower), "\n")
cat("acceleration Mean:", mean(acceleration), "\n")
cat("acceleration SD:", sd(acceleration), "\n")
```

(d) Now remove the 10th through 85th observations. What is the range, mean, and standard deviation of each predictor in the subset of the data that remains?
